---
title: "Telco Customer Churn Analysis"
author: "Victoire Migashane"
date: '`r Sys.Date()`'
output: pdf_document
---

# Question: What is causing Churn?

```{r setup, echo=TRUE, include=FALSE, comment= ''}
# Load necessary libraries
library(dplyr)      # For data manipulation
library(readxl)     # For reading Excel files
library(ggplot2)    # For data visualization
library(gridExtra)  # For arranging multiple plots

# Read the Telco customer churn dataset from an Excel file
df <- read_excel("./Telco_customer_churn_data.xlsx")  # Ensure the file path is correct

dim(df[df[["Churn Label"]] == "Yes",])[1]
```

## Data Dictionary
The data comes from Kaggle.com website: [Click here to get full data](https://www.kaggle.com/datasets/aqleemkhan/telco-customer-churn?resource=download)

| Feature            | Definition |
|---------------|------------------------------------|
| `CustomerID`       | Unique customer identifier. |
| `Count`            | Count of records for the customer (always 1). |
| `Country`          | Customer's country. |
| `State`            | Customer's state. |
| `City`             | Customer's city. |
| `Zip Code`         | Customer's postal code. |
| `Lat Long`         | Customer's latitude and longitude. |
| `Latitude`         | Customer's latitude. |
| `Longitude`        | Customer's longitude. |
| `Gender`           | Customer's gender (Male/Female). |
| `Senior Citizen`   | Senior citizen status (1 = Yes, 0 = No). |
| `Partner`          | Whether the customer has a partner (Yes/No). |
| `Dependents`       | Whether the customer has dependents (Yes/No). |
| `Tenure Months`    | Number of months with the company. |
| `Phone Service`    | Whether the customer has phone service (Yes/No). |
| `Multiple Lines`   | Whether the customer has multiple phone lines (Yes/No/No service). |
| `Internet Service` | Type of internet service (DSL, Fiber, None). |
| `Online Security`  | Whether the customer has online security (Yes/No/No service). |
| `Online Backup`    | Whether the customer has online backup (Yes/No/No service). |
| `Device Protection`| Whether the customer has device protection (Yes/No/No service). |
| `Tech Support`     | Whether the customer has tech support (Yes/No/No service). |
| `Streaming TV`     | Whether the customer has streaming TV (Yes/No/No service). |
| `Streaming Movies` | Whether the customer has streaming movies (Yes/No/No service). |
| `Contract`         | Type of contract (Month-to-month, 1 year, 2 years). |
| `Paperless Billing`| Whether the customer has paperless billing (Yes/No). |
| `Payment Method`   | Payment method (Bank transfer, Credit card, etc.). |
| `Monthly Charges`  | Monthly charge for the service. |
| `Total Charges`    | Total charges during the tenure. |
| `Churn Label`      | Whether the customer churned (Yes/No). |
| `Churn Value`      | Churn indicator (1 = Churned, 0 = Active). |
| `Churn Score`      | Churn prediction score.|
| `CLTV`             | Customer lifetime value (future revenue estimate). |
| `Churn Reason`     | Reason for churn (if applicable). |

# Wrangle

## Dataset Summary
```{r understand, echo=TRUE, comment=""}
# Check the dimensions of the dataset
print(dim(df))  # The dataset contains 7,043 rows and 33 columns

# Identify numeric vs. non-numeric (categorical) columns
numeric_cols <- df %>% select(where(is.numeric))  # Select numeric columns
numeric_cols %>% ncol()  # 10 columns are numeric

non_numeric_cols <- df %>% select(!where(is.numeric))  # Select non-numeric columns
non_numeric_cols %>% ncol()  # 23 columns are non-numeric (categorical/text)

# Display the column names of the dataset
colnames(df)

# Get descriptive statistics for numeric columns
summary(numeric_cols)

# Count the number of missing values in each column
colSums(is.na(df))  # 'Total Charges' has 11 missing values, 'Churn Reason' has 5,174 missing values

```
### **Observations**

- **Total Rows**: 7,043  
- **Total Columns**: 33  
- **Numeric Columns**: 10  
- **String (Object) Columns**: 23  
- **Missing Values**:  
  - `Total Charges`: 11 null values  
  - `Churn Reason`: 5,174 null values  
- **Numeric Data**: Descriptive statistics available  

## Preparation

### Data Cleaning
```{r prep, echo=TRUE, comment=""}
# Check the number of unique values in specific columns  
unique(df$Count)  # Only 1 unique value, so this column will be removed  
unique(df$Country)  # Only 1 unique value, so this column will be removed  
unique(df$State)  # Only 1 unique value, so this column will be removed  

# Define columns to remove  
removed_cols <- c("Count", "Country", "State", "Lat Long", "Latitude", "Longitude", "Churn Reason")  

# Remove the unwanted columns  
df <- df %>% select(-removed_cols)  

# Remove columns that still have missing values  
df <- df[, colSums(is.na(df)) == 0]  

# Check the new size of the dataframe  
df %>% dim()  # Displays the updated number of rows and columns  

# Replace spaces in column names with underscores for easier referencing  
colnames(df) <- gsub(" ", "_", colnames(df))  

# Identify numeric vs. non-numeric (categorical) columns  
numeric_cols <- df %>% select(where(is.numeric))  # Select numeric columns  
numeric_cols %>% ncol()  # 10 columns are numeric  

non_numeric_cols <- df %>% select(!where(is.numeric))  # Select non-numeric columns  
non_numeric_cols %>% ncol()  # 23 columns are categorical (string/objects)
```

### **Data Cleaning Summary**

**Dropped Columns:**

- **`count`**: Only had one value (1).
- **`Country`**: Only had one value (United States).
- **`State`**: Only had one value (California).
- **`Lat Long`**, **`Latitude`**, **`Longitude`**: Not analyzing location at the moment.
- **`Churn Reason`**: Had too many missing values.

**New Dataset Summary:**

- **Total Rows**: 7,043
- **Total Columns**: 25
- **Numeric Columns**: 6
- **Categorical Columns**: 19

## Visualize the columns to try and understand them better
The start by looking at the distributions of the 6 numerical comuns
```{r prep_numeric_viz, echo=TRUE, fig.width=8, fig.height=4}
# Look at the distribution of the numeric columns
for (col in colnames(numeric_cols)) {
  fig <- ggplot(df, aes_string(x = col)) + 
  geom_histogram() +
  labs(title = col) +
  theme(plot.title = element_text(hjust = 0.5)) + # center the title
  theme_minimal()
  
  print(fig)
}

```

### **Observations**

- **`Zip Code`**: Not very significant in the distribution.  
- **`Tenure Months`**: Displays a distribution where outcomes are more likely to occur at the extremes.  
- **`Monthly Charges`**: Appears to have a somewhat bell-shaped or normal distribution.  
- **`Churn Value`**: About a third of all customers have churned in this dataset.  
- **`Churn Score`**: Shows some characteristics of a normal distribution but is not fully normal.  
- **`CLTV`**: Shows some characteristics of a normal distribution but is not fully normal.

```{r prep_object_viz2, echo=TRUE, fig.width=8, fig.height=4}
for (col in colnames(non_numeric_cols)) {
  if (length(unique(df[[col]])) < 10) {
    fig <- ggplot(df, aes(x = .data[[col]])) + 
      geom_bar() +  # Use geom_bar() for categorical data
      labs(title = col) +
      theme(plot.title = element_text(hjust = 0.5)) + 
      theme_minimal()
    
    print(fig)
  }
}
```

### **Observations**

- **`Gender`**: Customers are almost evenly split between both genders.  
- **`Senior Citizen`**: The majority of customers are not senior citizens.  
- **`Phone Service`**: Most customers have phone service.  
- **`Multiple Lines`**: Most customers do not have multiple lines, but a significant portion does. Some have no phone service.  
- **`Internet Service`**: Most customers prefer fiber optic internet.  
- **`Online Security`**: Customers show little interest in online security services.  
- **`Online Backup`**: Low interest, possibly for similar reasons as online security.  
- **`Device Protection`**: Similar to online security, customers show low interest.  
- **`Tech Support`**: About half of the customers with internet service receive tech support.  
- **`Contract`**: The majority of customers are on a month-to-month contract.  
- **`Paperless Billing`**: Most customers prefer paperless billing.  
- **`Payment Method`**: Electronic checks are the most commonly used payment method.  
- **`Churn Label`**: About a third of all customers have churned in this dataset.  
 

# Exploritory Analysis

## Features of Interest dictionary

| Feature            | Definition |
|---------------|------------------------------------|
| `Churn Label`      | Whether the customer churned (Yes/No). |
| `Gender`           | Customer's gender (Male/Female). |
| `Churn Score`      | Churn prediction score.|
| `Senior Citizen`   | Senior citizen status (1 = Yes, 0 = No). |
| `Contract`         | Type of contract (Month-to-month, 1 year, 2 years). |
| `Paperless Billing`| Whether the customer has paperless billing (Yes/No). |
| `Payment Method`   | Payment method (Bank transfer, Credit card, etc.). |
| `Monthly Charges`  | Monthly charge for the service. |
| `CLTV`             | Customer lifetime value (future revenue estimate). |
| `Phone Service`    | Whether the customer has phone service (Yes/No). |
| `Multiple Lines`   | Whether the customer has multiple phone lines (Yes/No/No service). |
| `Internet Service` | Type of internet service (DSL, Fiber, None). |
| `Tech Support`     | Whether the customer has tech support (Yes/No/No service). |
| `Tenure Months`    | Number of months with the company. |

I will visualize each feature (y-axis) with my targe variable churn (x-axis) using box plot, violine plot, bar plot to full see the how each feature relate with my target variable.

```{r bivariate, echo=TRUE, fig.width=8, fig.height=4}
# Define numeric features to evaluate  
num_features <- c("Tenure_Months", "Monthly_Charges", "Churn_Score", "CLTV")  

# Create an empty list to store plots  
plot_list <- list()  

# Generate boxplot and violin plot for each numeric feature  
for (feature in num_features) {  
  p <- ggplot(df, aes(x = factor(Churn_Label), y = .data[[feature]])) +  
    geom_boxplot() +  
    geom_violin(alpha = 0.5, fill = "lightblue") +  
    labs(title = paste("Churn_Label vs", feature), x = "Churn_Label", y = feature) +  
    theme_minimal()  
  plot_list <- append(plot_list, list(p))  # Store each plot in the list  
}  

# Arrange and display two plots per page  
for (i in seq(1, length(plot_list), by = 2)) {  
  grid.arrange(plot_list[[i]], plot_list[[i + 1]], ncol = 2)  
}  
```

### **Observations**

- **Tenure Month vs Churn Label**:  
  Like we originally saw in the histogram, the `No` results are very balanced with a U-shape, showing outcomes happening at the extremes. However, the `Yes` category shows a large concentration of churn between 0 and 30 months as a Telco customer. This is significant because it could indicate that most customers churn before reaching three years with Telco. **This brings up the question: Why are they leaving the company so soon?**

- **Monthly Charges vs Churn Label**:  
  The focus of this plot is the violin plot, which provides insight into data density. Looking at the shape for `No`, we see that results are spread over a wide range of data points, and the box plot confirms this. The `Yes` category, however, shows a large concentration of data points at the higher end of monthly charges. This is very informative because it could suggest that **customers churn due to high monthly costs**.

- **Churn Score vs Churn Label**:  
  The churn score is assigned to every customer based on historical data and demographics, representing Telco’s prediction of the likelihood that a customer will churn. This plot shows how accurate Telco's predictions are, as there is little overlap between high and low churn scores. This suggests that **the higher the churn score, the higher the likelihood of churn**.

- **CLTV vs Churn Label**:  
  CLTV represents a customer's estimated lifetime revenue. It is also a prediction made by Telco's algorithms to understand the value of their customers. While there is not a significant difference between the two groups, one might argue that **customers who churn tend to have high lifetime revenue expectations from Telco**. However, the difference between medians is not substantial, so hypothesis testing could determine whether this difference is statistically significant.  

## Important Features

| Feature            | Definition |
|---------------|------------------------------------|
| `Churn Score`      | Churn prediction score.|
| `Monthly Charges`  | Monthly charge for the service. |
| `CLTV`             | Customer lifetime value (future revenue estimate). |
| `Tenure Months`    | Number of months with the company. |

```{r bivariate2, echo=TRUE, fig.width=8, fig.height=4}
# Create an empty list to store plots  
plot_list <- list()  

# Define categorical features to evaluate  
cat_features <- c("Gender", "Senior_Citizen", "Phone_Service", "Multiple_Lines", "Tech_Support", "Contract", "Paperless_Billing", "Payment_Method")  

# Generate bar plots for each categorical feature  
for (feature in cat_features) {  
  p <- ggplot(df, aes(x = factor(Churn_Label), fill = factor(.data[[feature]]))) +  
    geom_bar(position = "dodge") +  
    labs(title = paste("Churn_Label vs", feature), x = "Churn_Label", fill = feature) +  
    theme_minimal()  
  plot_list <- append(plot_list, list(p))  # Store each plot in the list  
}  

# Arrange and display two plots per page  
for (i in seq(1, length(plot_list), by = 2)) {  
  grid.arrange(plot_list[[i]], plot_list[[i + 1]], ncol = 2)  
}
```

### **Observations**

- **Counts of Churn Label vs Gender**:  
  There are no significant gender differences related to customer churn. **More customers stay with Telco than those who leave.**

- **Counts of Churn Label vs Senior Citizen**:  
  The majority of customers are not seniors. Non-senior citizens churn more than senior citizens, but this might be due to their larger presence in the dataset. Senior citizens churn at about the same volume as those who don’t. **I don’t think there is much significance here.**

- **Counts of Churn Label vs Phone Service**:  
  Only a small number of customers with phone service churn compared to those without it. This raises questions about customer satisfaction. **Are customers churning due to poor phone service coverage?**

- **Counts of Churn Label vs Multiple Lines**:  
  Customers who churn with multiple lines are about the same as those who churn with only a single line. **It doesn’t seem like the number of lines is a major factor in churn.**

- **Counts of Churn Label vs Tech Support**:  
  It is very clear that customers who churn have overwhelmingly received no technical support, while the opposite is true for those who stay. **Why didn’t these customers receive any technical support?**

- **Counts of Churn Label vs Contract**:  
  Month-to-month customers are churning at an extreme rate. **This could be a significant feature in predicting churn.**

- **Counts of Churn Label vs Paperless Billing**:  
  Paperless billing appears to play some role in customer churn, but its significance is unclear. **I think hypothesis testing would help a lot in understanding how important this feature is.**  

- **Counts of Churn Label vs Payment Method**:  
  Customers who use electronic checks churn at an extreme rate compared to every other group. **What is it about electronic checks that pushes customers away?**  

## Important Features

| Feature            | Definition |
|---------------|------------------------------------|
| `Contract`         | Type of contract (Month-to-month, 1 year, 2 years). |
| `Payment Method`   | Payment method (Bank transfer, Credit card, etc.). |
| `Phone Service`    | Whether the customer has phone service (Yes/No). |
| `Tech Support`     | Whether the customer has tech support (Yes/No/No service). |

```{r pressure, echo=TRUE, fig.width=8, fig.height=4}
# Boxplot of Monthly_Charges vs Churn_Label, colored by Internet_Service  
ggplot(df, aes(x = factor(Churn_Label), y = Monthly_Charges, color = factor(Contract))) +  
  geom_boxplot() +  
  labs(title = "Churn Label vs Monthly Charges by Contract",  
       x = "Churn Label", y = "Monthly Charges") +  
  theme_minimal()  

# Bar plot of Contract vs Churn_Label, faceted by Internet_Service  
ggplot(df, aes(x = factor(Churn_Label), fill = factor( Tech_Support))) +  
  geom_bar(position = "dodge") +  
  facet_wrap(~ Contract) +  
  labs(title = "Churn Label vs Contract by  Tech Support",  
       x = "Churn_Label", fill = "Tech Suppotr") +  
  theme_minimal()  

# Scatter plot of Monthly_Charges vs CLTV, colored by Churn_Label  
ggplot(df, aes(x = Churn_Score, y = CLTV, color = factor(Churn_Label))) +  
  geom_point(alpha = 0.6) +  
  labs(title = "Churn Score vs CLTV by Churn Label",  
       x = "Churn Score", y = "CLTV") +  
  theme_minimal()
```

### **Observations**

- **Churn Score vs CLTV by Churn Label**:  
  The data suggests that all customers pay roughly the same monthly charges, but longer tenure leads to higher lifetime revenue (CLTV) for Telco. A key concern is month-to-month customers who churn at a higher rate despite paying lower monthly charges collectively. This raises a possible **income issue**—  
  **Insight:** *These customers may not have high earnings, making retention more challenging.*

- **Churn Label vs Contract by Tech Support**:  
  Customers with technical support appear to churn less, suggesting that support services play a crucial role in customer satisfaction and retention.  
  **Takeaway:** *Providing better tech support could reduce churn rates.*

- **Churn Score vs CLTV by Churn Label**:  
  It’s unclear how Telco determines Churn Score and CLTV, but both seem to be **strong predictors of customer churn**. The relationship between expected lifetime revenue and churn score is highly correlated with churn behavior**.  
  **Next Step:** Understanding how these values are calculated could provide deeper insights into churn patterns and predictive accuracy.

## Conclusion

Below are the most predictive feature I have found for churn:

- Contract
- Payment Method
- Phone Service
- Tech Support
- Churn Score
- Monthly Charges
- CLTV
- Tenure Months

